---
title: "Presentation"
author: "Guy"
format: html
include-in-header: 
   - text: 
       \usepackage{tabularx}
       \usepackage[dvipsnames]{xcolor}
editor: visual
---

# Day 1: Data handling and Descriptive statistics

## 1. R coding refresher

### 1.1 R Math

Let's start with the fundamentals. To get acquainted with the R programming environment, begin with some basic computations. The R Console can also function as an interactive calculator.

```{r}
log(1) + sqrt(4)
1/3 + exp(1)
log(100, base = 10)
4^2 + 2
sum(4^2, 2)

```

The *ceiling()* function rounds a number upwards to its nearest integer, and the *floor()* function rounds a number downwards to its nearest integer, and returns the result:

```{r}
ceiling(1.4)
floor(1.4) 
```

### 1.2 R comments

You can use the comments before a line code

```{r}
#Print Good morning
"Good morning"
```

You can use it at the end of the line code

```{r}
"Good morning" #Print Good morning
```

You can also use multiple line comments

```{r}
# What is the value of
# square root of 3
print(sqrt(3))
```

### 1.3 Variables

To create variables in R you can do the following

```{r}
Surname <- "Alice"
Age <- 12
print(Surname)
Age
```

To concatenate multiple words you can use the function *paste()*

```{r}
text1 <- "Alice is"
text2 <- "beautiful"
paste(text1, text2)
```

How to assign a real value to a variable?

```{r}
x <- 2
y <- 3
x+y
```

Can we concatenate a text with a numeric?

```{r}
paste("mum", 6)
```

Can we assign multiple variables to a same value?

```{r}
x <- y <- z <- 16
x
y
z
```

### 1.4 R Data types

The *class()* function can be used to check the data types of a variable

```{r}
a <- 5L
class(a)

b <- 5.5
class(b)

c <- 1 + 1i
class(c)

x <- "Hi everyone"
class(x)

y <- FALSE
class(y)
```

Can we convert a from numeric to complex?

```{r}
x <- 2
y <- as.complex(x)
y
class(y)
```

### 1.5 R string

Can we count the number of characters in a string?

```{r}
str <- "Brilliant"
nchar(str)
```

Can we check if a character or a sequence of characters are present in a string?

```{r}
str <- "Good Morning!"

grepl("d", str)
grepl("s", str)
grepl("Good", str)

```

Can I write a text which contains word(s) in quotation marks?

```{r}
str <- "Have you heard the song \"Don't Stop Believing in yourself\"?"
cat(str)
```

### 1.6 R Booleans

When you compare two values, the expression is evaluated and R returns the logical answer(TRUE or False):

```{r}
x <- -1
y <- 0
x == y
```

### 1.6 The if statement

An "if statement" is created using the `if` keyword and is utilized to define a block of code that executes when a condition evaluates to TRUE.

```{r}
x <- 15
y <- 3

if (x > y) {
  print("x is greater than y")
} 
```

### 1.7 Else If

The `else if` keyword is R's way of saying "if the previous conditions were not true, then try this condition":

```{r}
a <- 10
b <- 25

if (b > a) {
  print("b is greater than a")
} else if (a == b) {
  print ("a and b are equal")
} 
```

### 1.8 If Else

The `else` keyword catches anything which isn't caught by the preceding conditions:

```{r}
a <- 10
b <- 15

if (b > a) {
  print("b is greater than a")
} else if (a == b) {
  print("a and b are equal")
} else {
  print("a is greater than b")
} 
```

You can also use `else` without `else if`:

```{r}
a <- 213
b <- 310

if (b > a) {
  print("b is greater than a")
} else {
  print("b is not greater than a")
} 
```

### 1.9 R While Loops

With the `while` loop we can execute a set of statements as long as a condition is TRUE. [Example]{style="color:red;"}: Print `i` as long as `i` is less than 6:

```{r}
i <- 1
while (i < 6) {
  print(i)
  i <- i + 1
} 
```

### 1.10 Break

With the `break` statement, we can stop the loop even if the while condition is TRUE. [Example]{style="color:red;"}: Exit the loop if `i` is equal to 4.

```{r}
i <- 1
while (i < 6) {
  print(i)
  i <- i + 1
  if (i == 4) {
    break
  }
} 
```

### 1.11 Next

With the `next` statement, we can skip an iteration without terminating the loop. [Example]{style="color:red;"}: Skip the value of 3.

```{r}
i <- 0
while (i < 6) {
  i <- i + 1
  if (i == 3) {
    next
  }
  print(i)
} 
```

### 1.12 R For Loop

A `for` loop is used for iterating over a sequence.

[Example1]{style="color:red;"}: Print the first 10 natural numbers.

```{r}
for (x in 0:9) {
  print(x)
} 
```

[Example2]{style="color:red;"}: Print every item in a list

```{r}
fruits <- list("apple", "banana", "cherry")

for (x in fruits) {
  print(x)
}
```

[Example3]{style="color:red;"}: Print every item in a vector

```{r}
vec <- c(1, 2, 3, 4, 5, 6)

for (x in vec) {
  print(x)
} 
```

### 1.13 R Functions

#### 1.13.1 Creating a function

To create a function, use the `function()` keyword:

```{r}
my_function <- function() {      # create a function with the name my_function
  print("Hello World!")
} 
```

#### 1.13.2 Call a Function

To call a function, use the function name followed by parenthesis, like **my_function()**:

```{r}
my_function <- function() {
  print("Hello World!")
}

my_function() # call the function named my_function 
```

#### 1.13.3 Arguments

Information can be passed into functions as arguments.

Arguments are specified after the function name, inside the parentheses. You can add as many arguments as you want, just separate them with a comma.

The following example has a function with one argument (fname). When the function is called, we pass along a first name, which is used inside the function to print the full name:

```{r}
my_function <- function(fname) {
  paste(fname, "Denis")
}

my_function("Lara")
my_function("Kirsten")
my_function("Frank") 
```

### 1.13.4 Number of Arguments

By default, a function must be called with the correct number of arguments. Meaning that if your function expects 2 arguments, you have to call the function with 2 arguments, not more, and not less:

```{r}
my_function <- function(fname, lname) {
  paste(fname, lname)
}

#my_function("Peter") 
my_function("Kirsten", "Denis") 

```

#### 1.13.5 Return Values

To let a function return a result, use the return() function:

```{r}
my_function <- function(x) {
  return (3 * x)
}
print(my_function(5))
```

## 2. Descriptive Statistics

### 2.1 Importing CSV file into R

For importing data CSV file into R, we can use the following code. When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
chap1data1 <-read.csv("chap1data1.csv", sep=",", header=T)
```

### 2.2 Visualize the data

```{r}
head(chap1data1)
tail(chap1data1)
```

### 2.3 Categorical variables

To prevent categorical variables from being treated as numerical data, we need to define them as factors.

```{r}
chap1data1$Male <- as.factor(chap1data1$Male)
chap1data1$Married<- as.factor(chap1data1$Married)
chap1data1$College <- as.factor(chap1data1$College)
chap1data1$Distance <- as.factor(chap1data1$Distance)
chap1data1$Treadmill <- as.factor(chap1data1$Treadmill)
chap1data1$Shoes <- as.factor(chap1data1$Shoes) 
```

### 2.4 Installing and loading required packages

To find out more of a specific package run the ??\[package name\]

```{r}
install.packages('RcmdrMisc')
library(RcmdrMisc)
```

### 2.5 Summary statistics

We can get summary statistics of a spesific variable, or all variables. The function numSummary can be used to get spesific statistics

mean: measure of centrality sd: standard deviation measure of distribution skewness: measure of distribution and normality kurtosis: indication of form of frequency distribution

```{r}
summary(chap1data1)
summary(chap1data1[,"Age"])
sapply(chap1data1,summary)

numSummary(chap1data1[,"Age"],statistics=c("mean","sd","skewness","kurtosis"))
```

### 2.6 Visualisation

Below is a list of different type of graphs used according to a given data type:

-   Bar chart (for categorical or discrete numerical data)
-   Line plot (similar to a bar chart, used to emphasize discreteness of the variable)
-   Histogram (for continuous data)
-   Pie chart (for categorical data measured on a nominal scale)
-   Scatter plot (to inspect the relationship between two variables)
-   Box plot (to inspect the distribution of a numerical variable measured on a continuous scale)

We will use two plotting systems in R in this presentation:

1.  Base plotting
2.  The ggplot system

Let us start with plotting the aforementioned list of different type of graphs with the R base plot

#### 2.6.1 Bars plot

```{r}
##Type of running
barplot(table(chap1data1$Type.of.running),main="Type of Running",ylab="Frequency",xlab="Type of Running")
#barplot(table(chap1data1[,"Type.of.running"]),main="Type of running",ylab="Frequency",xlab="Type of running",width=1,space=0)
```

#### 2.6.2 Line plot

```{r}
###Shoes
#chap1data1[,"Shoes"]

#plot(chap1data1[,"Shoes"],main="Type of Shoes",ylab="Frequency",xlab="Type of Shoes",col="red")
barplot(table(chap1data1[,"Shoes"]),main="Type of Shoes",ylab="Frequency",xlab="Type of Shoes",col="green", width =1, space =100)
```

#### 2.6.4 Histogram

```{r}
##Age
hist(chap1data1[,"Age"],main="Histogram for Age",xlab="Age")
```

#### 2.6.5 Pie chart

```{r}
###Type of running
pie(table(chap1data1[,"Type.of.running"]),main="Pie chart for types of running")
```

#### 2.6.5 Scatterplot

```{r}
## Age and Miles.per.week 
plot(x=chap1data1[,"Age"],y=chap1data1[,"Miles.per.week"], main="Age vs Miles per week",ylab="Miles per week",xlab="Age")
```

#### 2.6.6 Box plot

```{r}
## Miles.per.week
boxplot(chap1data1[,"Miles.per.week"],main="Miles per week Boxplot",ylab="Miles per week",xlab="")
```

#### 2.6.7 Box plot by groups

```{r}
## Miles.per.week by Shoes
boxplot(Miles.per.week~Shoes,data=chap1data1,main="Miles per week by Shoes Boxplot",ylab="Miles per week",xlab="Shoes")
```

To use the ggplot plotting system you need to install the package first and load it into R

```{r}
install.packages('ggplot2')
library(ggplot2)

#or do the following

#In the miscellaneous window, click on Packages.
#Click on Install, then type “ggplot2” beneath Packages (separate multiple with space or comma)
#Make sure Install dependencies is selected, then click on Install
#After installation, load the package into R: package(ggplot2)

```

Construct the same graphs (except the line plot) with ggplot plotting system.

#### 2.6.8 Bar plot -- Shoes

```{r}
barplot <- ggplot(data=chap1data1,aes(x=Shoes)) +
geom_bar(fill="gray",color="black",width=.5)
barplot + ggtitle("Barplot of Types of Shoes") + labs(x="Types of
Shoes",y="Frequency") +
theme(plot.title=element_text(size=20),axis.title=element_text(size=
14))
```

### 2.6.9 Bar plot by groups -- Shoes by Male

```{r}
barplot2 <- ggplot(data=chap1data1,aes(x=Shoes,y=Miles.per.week))
barplot2 + geom_bar(aes(fill=Male),
position=position_dodge(),stat="identity") +
scale_fill_manual(values=c("#CC6666","#66CC99"))
```

#### 2.6.10 Histogram -- Age

```{r}
histo <- ggplot(chap1data1,aes(x=Age))
histo + geom_histogram(binwidth=6)
```

#### 2.6.11 Pie chart -- Shoes

```{r}
pie <- ggplot(data=chap1data1,aes(x=factor(1),fill=Shoes))+
geom_bar(width = 1)
pie + coord_polar("y")
```

#### 2.6.12 Scatter plot -- Runs.per.week vs Age

```{r}
ggplot(data=chap1data1,aes(y=Runs.per.week,x=Age)) +
geom_point(shape=3,color="darkblue")
```

#### 2.6.13 Scatter plot -- Runs.per.week vs Age by Distance

```{r}
scatter <-
ggplot(data=chap1data1,aes(y=Runs.per.week,x=Age,color=Distance))
scatter + geom_point(size=2) +
scale_color_manual(values=c("#CC6666","#66CC99"))
```

#### 2.6.14 Scatter plot -- Runs.per.week vs Age by Distance (with trend lines)

```{r}
scatter2 <-
ggplot(data=chap1data1,aes(y=Runs.per.week,x=Age,color=Distance))
scatter2 + geom_point(size=2) +
scale_color_manual(values=c("#CC6666","#66CC99")) +
geom_smooth(method=lm,se=F)
```

#### 2.6.15 Box plot -- Miles.per.week by Distance

```{r}
ggplot(chap1data1,aes(factor(Distance),Miles.per.week)) +
geom_boxplot()
```

#### 2.6.15 Exercises

##### Exercise 1

For the data in C&S Example 2.1 (p. 11) (chap1exer1.csv), calculate the following statistics regarding the sample plant height.

-   Mean
-   Median
-   Variance
-   Standard deviation
-   Standard error of the mean
-   Coefficient of variation

You can also try to import the data with the following function: "matrix(c(14.8,15.2,17.4,11.6,12.5),ncol=1)"

##### Exercise 2

See if you can import the data below with the following function: data.frame(Yield=c(8.1,8.7,9.2,7.8,8.4,9.4))

-   Report the mean, median and mode.
-   Report the variance and standard deviation.
-   Produce a line plot.

#### Exercise 3

The data in C&S Example 3.2 (p. 18) (chap1exer3.csv), presents the yields (g) of small equal-sized plots of barley. Determine the following:

-   mean

-   mimimum and maximum

-   standard deviation

-   range

-   first and third quartiles

-   interquartile range and also

-   produce a histogram

-   produce a box plot.
