---
title: "RM"
author: "Cecile Bester"
date: "2024-04-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Required packages

```{r}
library(RcmdrMisc)
library(ggplot2)
library(nlme)
library(emmeans)
```

## --\> Chapter9 - Examples

#Example 1

```{r}
chap9data1 <- read.csv("chap9data1.csv")
chap9data1$Food <- as.factor(chap9data1$Food)
chap9data1$Calf <- as.factor(chap9data1$Calf)
chap9data1$Month <- as.factor(chap9data1$Month)
```

#fit model

```{r}
chap9data1.model1 <- lme(Weight~Month*Food,random=~1|Calf,data=chap9data1)
anova(chap9data1.model1)

chap9data1.model2 <- aov(Weight~Food*Month + Error(Calf),data=chap9data1)
summary(chap9data1.model2)
```

## Interpretation of the ANOVA Results

The analysis of variance (ANOVA) was conducted using a linear mixed-effects model (`lme`) to understand the effects of `Month`, `Food`, and their interaction (`Month:Food`) on the `Weight` of calves. Here is the interpretation of the ANOVA table generated:

### 1. Intercept:

-   **F-value:** 465.2000\
-   **p-value:** `<.0001`\
-   This indicates that the intercept is highly significant. In mixed-effects models, the intercept represents the baseline level of the dependent variable (`Weight`) when all predictors are at their reference level. A highly significant intercept suggests that the mean `Weight` of the calves is significantly different from zero.

### 2. Month:

-   **Degrees of Freedom (numDF):** 3\
-   **F-value:** 15.6277\
-   **p-value:** `<.0001`\
-   The variable `Month` has a highly significant effect on `Weight` (p-value \< 0.0001). This suggests that the average `Weight` of the calves changes significantly across different months.

### 3. Food:

-   **Degrees of Freedom (numDF):** 1\
-   **F-value:** 1.4088\
-   **p-value:** 0.2693\
-   The variable `Food` does not have a statistically significant effect on `Weight` (p-value = 0.2693). This indicates that, after accounting for other variables in the model, different types of food do not significantly affect the calves' weight.

### 4. Interaction Between Month and Food (Month:Food):

-   **Degrees of Freedom (numDF):** 3\
-   **F-value:** 0.8711\
-   **p-value:** 0.4697\
-   The interaction between `Month` and `Food` is not statistically significant (p-value = 0.4697). This suggests that the effect of `Food` on `Weight` does not vary significantly across different months.

## Conclusion:

-   The results indicate that the calves' weight varies significantly over the months (`Month` is significant), but there is no significant effect of `Food` or the interaction between `Month` and `Food` on the weight. Therefore, month-to-month changes are the main driver of differences in weight, rather than the type of food or any combined effect of the month and food type. """

## Interpretation of the ANOVA Summary

The analysis of variance (ANOVA) was conducted using the `aov` function with an `Error` term for the random effect of `Calf`. This model evaluates the effects of `Food`, `Month`, and their interaction (`Food:Month`) on the `Weight` of calves while accounting for the random effect of `Calf`. Here is the interpretation of the ANOVA summary:

### 1. Error: Calf (Random Effect)

-   This section represents the variance attributed to differences between calves (treated as a random effect).
-   **Food:**
    -   **Degrees of Freedom (Df):** 1\
    -   **Sum of Squares (Sum Sq):** 409.6\
    -   **Mean Square (Mean Sq):** 409.6\
    -   **F-value:** 1.409\
    -   **p-value (Pr(\>F)):** 0.269\
    -   The effect of `Food` is not statistically significant (p-value = 0.269). This suggests that differences in the type of food do not significantly contribute to variations in weight among different calves.

### 2. Error: Within (Within-Groups Effects)

-   This section shows the variance explained by the fixed effects (`Month` and `Food:Month`) within the groups (calves).
-   **Month:**
    -   **Degrees of Freedom (Df):** 3\
    -   **Sum of Squares (Sum Sq):** 1036.9\
    -   **Mean Square (Mean Sq):** 345.6\
    -   **F-value:** 15.628\
    -   **p-value (Pr(\>F)):** 7.58e-06 (very significant, denoted by \*\*\*)\
    -   The effect of `Month` on `Weight` is highly significant (p-value \< 0.0001). This indicates that the weight of calves changes significantly across different months.
-   **Food:Month Interaction:**
    -   **Degrees of Freedom (Df):** 3\
    -   **Sum of Squares (Sum Sq):** 57.8\
    -   **Mean Square (Mean Sq):** 19.3\
    -   **F-value:** 0.871\
    -   **p-value (Pr(\>F)):** 0.47\
    -   The interaction between `Food` and `Month` is not statistically significant (p-value = 0.47). This suggests that the effect of `Food` on `Weight` does not vary significantly over different months.

### 3. Residuals

-   Represents the unexplained variance within the groups after accounting for the fixed effects (`Month` and `Food:Month`).
-   **Degrees of Freedom (Df):** 24\
-   **Sum of Squares (Sum Sq):** 530.8\
-   **Mean Square (Mean Sq):** 22.1

## Conclusion

-   The weight of calves varies significantly across different months, indicating a strong effect of `Month`.
-   There is no significant effect of `Food` alone or the interaction between `Food` and `Month` on the weight of calves.
-   The differences between individual calves (random effect of `Calf`) and other unexplained factors account for the remaining variance.

# Multiple comparisons

```{r}
# Calculate estimated marginal means for the model
emmeans_result <- emmeans(chap9data1.model1, ~ Month)

# Display the estimated marginal means
summary(emmeans_result)

# Conduct pairwise comparisons
pairs(emmeans_result)
```
