---
title: "SLR"
format: html
editor: visual
---

# I- Introduction

(See Biometry 721-821 workbook (2024) page 43)

# II- The Basics

## Load required pacakges

```{r}
library(RcmdrMisc)
library(ggplot2)
```

## Importing data

```{r}
chap6data1 <- read.csv("chap6data1.csv", sep=",", header=T)
chap6data2 <- read.csv("chap6data2.csv", sep=",", header=T)
View(chap6data1)
View(chap6data2)
```

#Example 1 #visual presentation of data

```{r}
chap6data1.plot1 <- ggplot(data=chap6data1,aes(y=MetabolicRate,x=BodyWeight)) 
chap6data1.plot1 + geom_point(size=2) 
chap6data1.plot1 + geom_point(size=2)+ geom_smooth(method=lm,se=F)
```

## Fit the model

```{r}
chap6data1.model1 <- lm(MetabolicRate~BodyWeight,data=chap6data1)
summary(chap6data1.model1)
anova(chap6data1.model1)
```

The formula of the fitted model is

$$Y = 811.2267 + 7.0595 X$$

## Interpretation of the parameter estimates

### Intercept

The average metabolic rate when body weight is zero is approximately 811.2267 units.

### BodyWeight

For each additional unit of body weight, the metabolic rate increases by about 7.0595 units.

### Determine whether greater weight is associated with greater metabolism.

Given the positive coefficient ($7.0595$) for BodyWeight, we can conclude that greater weight is indeed associated with greater metabolism. This positive relationship suggests that as body weight increases, metabolic rate also increases.

# III- Diagnostic checks

## III-1- Determine and interpret the coefficient of determination of the metabolism data

Since Multiple R-squared($R^{2}$) is equal to $0.5539$, then approximately $55.39\%$ of the variability in Metabolic Rate can be explained by Body weight.

### Model diagnostics

To assess the influence of individual data points on the linear regression model `chap6data1.model1`, we use influence measures and studentized residuals.

```{r}
influence.measures(chap6data1.model1)
matrix(rstudent(chap6data1.model1),ncol=1) #Studentized Residuals
```

#### Influence Measures

The `influence.measures()` function in R provides a comprehensive set of influence diagnostics, including:

-   **DFBETAS**: Measures the effect of deleting each observation on the estimated coefficients.
-   **DFFITS**: Measures the effect of deleting each observation on the fitted values.
-   **COVRATIO**: Measures the effect of deleting each observation on the covariance matrix of the estimated coefficients.
-   **Cook's Distance**: Measures the influence of deleting each observation on the estimated regression coefficients.
-   **Hat values (Leverage)**: Measures the influence of each observation on the fitted values.

##### Interpretation

-   **DFBETAS**: Values close to 0 indicate little influence. Large absolute values (e.g., \> 1 for small datasets) suggest influential observations.
-   **DFFITS**: Values larger than 1 or $\frac{2\sqrt{k+1}}{n}$ (where $k$ is the number of predictors and $n$ number of observations) suggest influential observations.
-   **COVRATIO**: Values far from 1 indicate influential observations affecting the covariance matrix.
-   **Cook's Distance**: Values larger than 1 indicate highly influential observations.
-   **Hat values (Leverage)**: Values larger than $\frac{2k}{n}$ suggest high leverage points.

#### Studentized Residuals

Studentized residuals are residuals divided by an estimate of their standard deviation. They help identify outliers and are calculated as: $$r_{i} = \frac{\epsilon_{i}}{s(\epsilon_{i})}$$, where where $\epsilon_{i}$ is the residual for observation $i$ and $s(\epsilon_{i})$ is its standard deviation.

##### Interpretation

If $\vert r_{i} \vert > 2$, then there is potential outliers.



Use the following code to add the diagnostic measures and the fitted values to the original data set.
```{r}
within(chap6data1, {
FittedValues <- fitted(chap6data1.model1)
  Residuals <- residuals(chap6data1.model1)
  StudentizedResiduals <- rstudent(chap6data1.model1)
  HatValues <- hatvalues(chap6data1.model1)
  CooksDistance <- cooks.distance(chap6data1.model1)
  ObsNumber <- 1:nrow(chap6data1)
})

#head(chap6data1)
```

### Creating Influence.Cutoffs function 
```{r}
influence.cutoffs <- function(model){
  p <- length(model$coefficients)
  n <- length(model$residuals)
  
  DFBETAS <- 2/(n^0.5)
  DFFITS <- 2*(p/n)^0.5
  COVRATIO.lwr <- 1-3*p/n
  COVRATIO.upr <- 1+3*p/n
  cook.D <- 4/n
  HATDIAG <- 2*p/n
  
  list(DFBETAS=DFBETAS,DFFITS=DFFITS,COVRATIO.lwr=COVRATIO.lwr,COVRATIO.upr=COVRATIO.upr,cook.D=cook.D,HATDIAG=HATDIAG)
}

influence.cutoffs(chap6data1.model1)
```

#Assess assumptions: normality and homoscedasticity
```{r}
par(mfrow=c(2,2))
base::plot(chap6data1.model1)
par(mfrow=c(1,1))
shapiro.test(chap6data1.model1$residuals)
```

### Prediction
```{r}
chap6data1.pred <- data.frame(BodyWeight=c(70,100))
predict(chap6data1.model1,newdata=chap6data1.pred,interval="confidence")
predict(chap6data1.model1,newdata=chap6data1.pred,interval="prediction")
chap6data1.pred2 <- cbind(chap6data1.pred,predict(chap6data1.model1,newdata=chap6data1.pred,interval="confidence"))

chap6data1.predictions <- cbind(chap6data1,predict(chap6data1.model1,newdata=chap6data1.pred, interval="prediction"))
print(chap6data1.predictions)
g <- ggplot(chap6data1.predictions,aes(x=BodyWeight,y=MetabolicRate)) + geom_point() + geom_line(aes(y = fit),colour = "blue", size = 1) + stat_smooth(method=lm) + geom_ribbon(aes(ymin = lwr, ymax = upr),  fill = "blue", alpha = 0.4)
g + geom_point(data=chap6data1.pred2,aes(x=BodyWeight,y=fit),col="red")

```


# IV- Exercises

Exercise 1

With the data from C&S Example 7.1, p. 64 (chap6exer1.csv): Perform a simple linear regression of
Area (ð‘Œ) on Stem length (ð‘‹) and give the equation of the best-fitting regression line, as well as the
coefficient of determination (ð‘…2 ). Interpret the ð‘…2 value. Also test the assumptions underlying this
regression model, and clearly state your conclusions with regard to the validity of the fitted model.


Exercise 2

With the data from C&S Example 7.2, p. 69 (chap6exer2.csv): Perform a simple linear regression of
wheat yield (ð‘Œ) on nitrogen fertilizer (ð‘‹) and give the equation of the best-fitting regression line. Predict the wheat yield for a nitrogen fertilizer level of ð‘‹ = 80kg nitrogen per hectare (also include the confidence and prediction intervals in your answer, and clearly interpret).

Exercise 3

Do C&S Exercise 7.2 (p. 84). The data is in the file chap6exer3.csv. A study was undertaken to find
out if tree diameter measurements 1.5m above ground level can be used to predict heights for a certain
species. The measurements of 12 trees are shown in the csv file. Complete a thorough regression
analysis (significance, R-squared, assumptions).
