---
title: "MLR"
format: html
editor: visual
---

# Load required packages

```{r}
library(RcmdrMisc)
library(ggplot2)
library(car)
library(MASS)
```

## Importing data

```{r}
chap7data2 <- read.csv("chap7data2.csv", sep=",", header=T)
chap7data3 <- read.csv("chap7data3.csv", sep=",", header=T)
View(chap7data2)
View(chap7data3)
```

#scatter plot

```{r}
# Scatter plots of 'Detox' against each enzyme
ggplot(chap7data2, aes(x = Enzyme1, y = Detox)) + geom_point() + ggtitle("Detox vs Enzyme1") +
  theme_minimal()
ggplot(chap7data2, aes(x = Enzyme2, y = Detox)) + geom_point() + ggtitle("Detox vs Enzyme2") +
  theme_minimal()
ggplot(chap7data2, aes(x = Enzyme3, y = Detox)) + geom_point() + ggtitle("Detox vs Enzyme3") +
  theme_minimal()
ggplot(chap7data2, aes(x = Enzyme4, y = Detox)) + geom_point() + ggtitle("Detox vs Enzyme4") +
  theme_minimal()
ggplot(chap7data2, aes(x = Enzyme5, y = Detox)) + geom_point() + ggtitle("Detox vs Enzyme5") +
  theme_minimal()
```

#or

```{r}
plot(chap7data2)
```

#or

```{r}
scatterplotMatrix(~Detox + Enzyme1 + Enzyme2 + Enzyme3 + Enzyme4 + Enzyme5,
                  reg.line = lm,
                  smooth = FALSE,
                  spread = FALSE,
                  span = 0.5,
                  ellipse = FALSE,
                  levels = c(0.5, 0.9),
                  id.n = 0,
                  diagonal = 'density',
                  data = chap7data2)
```

## Interpretation

This document provides an interpretation of the scatterplot matrix that includes the response variable (`Detox`) and explanatory variables (`Enzyme1`, `Enzyme2`, `Enzyme3`, `Enzyme4`, and `Enzyme5`). The scatterplot matrix is used to visualize the relationships between these variables and identify potential correlations.

### 1. Relationships Between `Detox` and Each Enzyme

-   **`Detox` vs. `Enzyme1`**:\
    There is a positive linear relationship, as indicated by the upward trend in the scatter plot. The regression line is slightly upward, and the points are moderately scattered around it, suggesting a moderate positive correlation.

-   **`Detox` vs. `Enzyme2`**:\
    The scatter plot shows a very weak relationship with a mostly horizontal regression line. The points are widely scattered without any discernible pattern, suggesting little to no linear relationship between `Detox` and `Enzyme2`.

-   **`Detox` vs. `Enzyme3`**:\
    The plot shows a positive linear relationship, as evidenced by the upward trend and a moderately steep regression line. Points are somewhat dispersed, but there is a noticeable trend suggesting a moderate positive correlation.

-   **`Detox` vs. `Enzyme4`**:\
    There appears to be a slight negative linear relationship. The regression line slopes downward, but the points are widely scattered. This suggests a weak negative correlation between `Detox` and `Enzyme4`.

-   **`Detox` vs. `Enzyme5`**:\
    The scatter plot indicates a weak relationship with a relatively flat regression line. The points are fairly scattered, with no clear trend, suggesting little to no linear correlation between `Detox` and `Enzyme5`.

### 2. Relationships Among Explanatory Variables (`Enzyme1` to `Enzyme5`)

-   **`Enzyme1` vs. `Enzyme2`**:\
    There is a noticeable positive linear relationship with an upward trend. The regression line is moderately steep, and points are clustered around it, indicating a moderate positive correlation.

-   **`Enzyme1` vs. `Enzyme3`, `Enzyme4`, `Enzyme5`**:\
    All these plots show relatively moderate positive linear relationships, with upward trends and points moderately clustered around the regression lines.

-   **`Enzyme2` vs. `Enzyme3`**:\
    The relationship appears to be non-linear, with points forming a curved pattern. This suggests a non-linear relationship between `Enzyme2` and `Enzyme3`.

-   **`Enzyme2` vs. `Enzyme4`, `Enzyme5`**:\
    These plots show no clear pattern, suggesting little to no linear relationship between `Enzyme2` and either `Enzyme4` or `Enzyme5`.

-   **`Enzyme3` vs. `Enzyme4`**:\
    There seems to be a positive relationship with an upward trend in the scatter plot, though the points are somewhat scattered, indicating a moderate positive correlation.

-   **`Enzyme3` vs. `Enzyme5`**:\
    Shows a weak or no discernible relationship, with the points widely scattered around a flat regression line.

-   **`Enzyme4` vs. `Enzyme5`**:\
    No clear relationship is observed here, as the points are scattered without any noticeable pattern.

### 3. Diagonal Plots (Density Plots)

-   The diagonal plots display the distribution of each variable. For example:
    -   **`Detox`** appears to have a somewhat skewed distribution, possibly slightly left-skewed.
    -   **`Enzyme1`** and **`Enzyme3`** appear to be normally distributed with slight variation.
    -   **`Enzyme2`** shows a highly skewed distribution with a long tail.
    -   **`Enzyme4`** and **`Enzyme5`** show more uniform distributions with less obvious skewness.

### Conclusion

-   `Detox` has a moderate positive relationship with `Enzyme1` and `Enzyme3`, suggesting these enzymes might be more influential in predicting `Detox` levels.
-   `Enzyme2` and `Enzyme5` appear to have little to no linear relationship with `Detox`.
-   The relationships among explanatory variables vary, with some moderate positive linear correlations (like `Enzyme1` with `Enzyme2`), and some non-linear or weak relationships.

# Multiple linear regression

```{r}
chap7data2.model1 <- lm(Detox~.,data=chap7data2)
summary(chap7data2.model1)
```

## #the allSubsets() function

```{r}
allSubsets <- function(data,y.name="Y",perf.measure=c("adj.r.squared","r.squared","AIC","perc.error")){
  
  Cols <- names(data)
  Cols <- Cols[! Cols %in% y.name]
  n <- length(Cols)
  
  id <- unlist(
    lapply(1:n,
           function(i)combn(1:n,i,simplify=F)
    )
    ,recursive=F)
  
  Formulas <- sapply(id,function(i)
    paste(y.name,"~",paste(Cols[i],collapse="+"))
  )
  
  result.mat <- matrix(0,nrow=length(Formulas),6)
  result.mat[,1] <- Formulas
  
  #get all adjusted R2 values
  for(i in 1:length(Formulas)){result.mat[i,2] <- summary(lm(as.formula(Formulas[i]),
                                                             data=data))$adj.r.squared}
  #get all R2 values
  for(i in 1:length(Formulas)){result.mat[i,3] <- summary(lm(as.formula(Formulas[i]),
                                                             data=data))$r.squared}
  #get all AIC's
  
  for(i in 1:length(Formulas)){result.mat[i,4] <- AIC(lm(as.formula(Formulas[i]),
                                                         data=data))}
  #get all sigma's
  
  for(i in 1:length(Formulas)){result.mat[i,5] <- summary(lm(as.formula(Formulas[i]),
                                                             data=data))$sigma/mean(data[,y.name])}
  #get all mallows cp's
  
  for(i in 1:length(Formulas)){result.mat[i,6] <- mallows.cp(data=data,model=lm(as.formula(Formulas[i]),
                                                                                data=data),y.name=y.name)}
  
  
  
  colnames(result.mat) <- c("Model","adj.r.squared","r.squared","AIC","perc.error","mallows.cp")
  

  
  final.output <- data.frame(result.mat[order(result.mat[,perf.measure],decreasing=T),])
  
  return(final.output)
  
}

```

## calculating the Mallow's Cp statistic (function)

```{r}
mallows.cp <- function(data,model,y.name)
{
  
  n <- nrow(data)
  p <- length(model$coefficients)
  
  mean.sq.error <- summary(lm(as.formula(paste(y.name,"~.")),data=data))$sigma^2
  
  sse <- summary(model)$sigma^2*summary(model)$df[2]
  
  cp <- sse/mean.sq.error-(n-2*p)
  cp/100
}
```

# Variable selection

```{r}
stepwise(chap7data2.model1, direction='backward/forward', criterion='AIC')
#or
allSubsets(chap7data2,y.name="Detox",perf.measure="mallows.cp")
```

## Interpretation 1

The stepwise regression analysis identified that the best model for predicting `Detox` includes the variables `Enzyme1`, `Enzyme4`, and `Enzyme5`. This model was chosen because it has the lowest Akaike Information Criterion (AIC) value (64.86), indicating the best balance between model complexity and fit.

-   **Enzyme1** and **Enzyme4** have small positive effects on `Detox`, while **Enzyme5** has a larger positive effect.
-   The variables `Enzyme2` and `Enzyme3` were excluded from the final model as they did not significantly improve the model's fit.

This suggests that `Enzyme1`, `Enzyme4`, and `Enzyme5` are the most important predictors of `Detox` in this dataset.

## Interpretation 2

Based on the regression analysis and the adjusted R-squared values for various models, the following conclusions can be drawn:

1.  **Best Model**: The model `Detox ~ Enzyme1 + Enzyme2 + Enzyme3` has the highest adjusted R-squared value of 0.6823. This indicates that it explains about 68.23% of the variability in `Detox` and is the best model among all those considered in terms of fit.

2.  **Other Notable Models**:

    -   The models `Detox ~ Enzyme1 + Enzyme2 + Enzyme5` (Adjusted R-squared: 0.6687) and `Detox ~ Enzyme1 + Enzyme2 + Enzyme4` (Adjusted R-squared: 0.6582) also have high adjusted R-squared values. These models provide a good fit and are close in performance to the best model.

3.  **Poorly Performing Models**:

    -   Models with adjusted R-squared values close to or below 0, such as `Detox ~ Enzyme5` (-0.1217) and `Detox ~ Enzyme3` (0.3014), indicate a poor fit. These models do not explain much of the variability in `Detox` and are not suitable for prediction.

4.  **Model Selection**:

    -   The top models to consider for predicting `Detox` are:
        -   `Detox ~ Enzyme1 + Enzyme2 + Enzyme3` (Adjusted R-squared: 0.6823)
        -   `Detox ~ Enzyme1 + Enzyme2 + Enzyme5` (Adjusted R-squared: 0.6687)
        -   `Detox ~ Enzyme1 + Enzyme2 + Enzyme4` (Adjusted R-squared: 0.6582)
    -   These models provide a good balance between complexity (number of variables) and explanatory power.

### Conclusion:

The best model based on adjusted R-squared is `Detox ~ Enzyme1 + Enzyme2 + Enzyme3`, which explains 68.23% of the variance in `Detox`. Models with fewer variables generally have lower adjusted R-squared values and may not provide sufficient explanatory power, while models with negative adjusted R-squared values should be discarded as they do not fit the data well.



# Another example with data 3

```{r}
chap7data3.model1 <- lm(Y~.,data=chap7data3)
#chap7data3.model1 <- lm(Y~X1+X2+X3+X4,data=chap7data3)
summary(chap7data3.model1)

vif(chap7data3.model1)

base::plot(chap7data3)

stepwise(chap7data3.model1, direction='backward', criterion='AIC')
stepwise(chap7data3.model1, direction='forward', criterion='AIC')
stepwise(chap7data3.model1, direction='backward/forward', criterion='AIC')

allSubsets(chap7data3,y.name="Y",perf.measure="adj.r.squared")

chap7data3.model2 <- lm(Y ~ X1 + X2 + X4,data=chap7data3)
summary(chap7data3.model2)
vif(chap7data3.model2)

chap7data3.model3 <- lm(Y ~ X1 + X2,data=chap7data3)
summary(chap7data3.model3)
vif(chap7data3.model3)
```

#Assess normality

```{r}
shapiro.test(chap7data3.model3$residuals)
```

#Assess homoscedasticity

```{r}
plot(y=chap7data3.model3$residuals,x=chap7data3.model3$fitted.values,main="Residuals vs Yhat",ylab="residuals",xlab="Yhat")
```


# Exercises 
Do all the exercices in page 58 on document "Biometry 721/821 Workbook(2024).pdf"